Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Final proofreading before given to Thorsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2915 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c299d00ef83671e7ca69229fc32e04fbb01483ea

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link functions must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of statistical inference.
In addition, the employed tree structure allows to capture non-additive effects while forests 
enable the modeling of smooth changes.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and further extensions \citep{Larraondo+Inza+Lozano:2018} do not only
allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed for probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes, e.g., in air traffic management as considered in this study.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a narrow valley within the European Alps.
Topography channels the wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements (lagged by one hour) and their derivatives, such as 
wind direction and wind speed at the airport or spatial and temporal differences.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates used for splitting
  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.
  Note that in the meteorological context wind direction is defined on the 
  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind 
regimes and are consequently associated with a very low estimated concentration 
parameter~$\kappa$, i.e. a high estimated variance. 
In terms of covariates, the lagged wind direction (``persistence'') is mostly 
responsible for distinguishing the broad range of wind regimes listed above while
the pressure gradients and wind speed separate the data into subgroups with high 
vs.\ low precision. 
A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.

The remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular regression trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or \citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, (distributional) regression models are of high interest for all types
of data, including circular data.

@@

However, all circular additive models incorporate the previously discussed difficulties induced by

@@

An alternative to additive models, which can deal or even 
avoid the mentioned difficulties in circular regression modeling, are 
tree-based models. For the particular case of circular data, the methodology 
of circular regression trees and forests is presented in the following section.

@@

\subsection{Data}
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airport for Innsbruck and Vienna on an
hourly temporal resolution. As predictor variables, we use temporal information
and $1$-hourly resolved $10$\,min mean observations of various meteorological quantities such as 
wind direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within an area of about XXX\,km$^2$ are used. A topographical overview of
the airports and their surrounding areas with the station sites employed in
this study is provided in Figure~XXX. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

@@

    using the $31$~days centered around the day of interest over all available years in the training data.

@@

    the regime changes takes place in the response variable during the forecast

@@

    maximum of $50$ classes (argument \code{nmax = c("yx" = Inf, "z" = 50)}).

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular regression tree and the persistency model; except for the $3$\,h~forecast 
at Innsbruck, where the persistency model is outperformed by all others. 
In comparison to the circular regression tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores of wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains only a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift, 
e.g., from up-valley to down-valley. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a change of the wind regime seem to 
lead to a higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

@@

\section{Tree algorithm}
\label{app:algorithm}
This section provides a more detailed overview on the permutation-test-based CTree algorithm 
\citep{Hothorn+Hornik+Zeileis:2006}, 
specifically for circular data as applied to build the circular regression trees and forests 
in the presented case study. 
Alternatively, also the MOB algorithm, which is based on M-fluctuation tests, can be employed 
\citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.

@@

To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic
$c$ is employed, for example in a quadratic form
\begin{equation}
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
\end{equation}
where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $T_l$,
as derived by \cite{Strasser+Weber:1999} and used to standardize $t_l$, and $\Sigma_l^+$ 
is the Moore-Penrose inverse of $\Sigma_l$.
As an alternative, also a maximum form ($c_{\text{max}}$) can be considered such that
the maximum of the absolute values of the standardized linear statistic is returned.

>>>>

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link functions must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of statistical inference.
In addition, the employed tree structure allows to capture non-additive effects while forests 
enable the modeling of smooth changes.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach does not only
allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed for probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes, e.g., in air traffic management as considered in this study.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a narrow valley within the European Alps.
Topography channels the wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements (lagged by one hour) and their derivatives, such as 
wind direction and wind speed at the airport or spatial and temporal differences.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates selected for splitting
  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.
  Note that in the meteorological context wind direction is defined on the 
  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind 
regimes and are consequently associated with a very low estimated concentration 
parameter~$\hat{\kappa}$, i.e., a high estimated variance. 
In terms of covariates, the lagged wind direction (``persistence'') is mostly 
responsible for distinguishing the broad range of wind regimes listed above while
the pressure gradients and wind speed separate the data into subgroups with high 
vs.\ low precision. 
A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.

The remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution, and circular regression
models will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular regression trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and \citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the empirical histogram along with 
the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, (distributional) regression models are of high interest for all types
of data, including circular data.

@@

Nevertheless, all circular additive models incorporate the previously discussed difficulties induced by

@@

An alternative to additive models, which can deal or even 
avoid the mentioned difficulties in circular regression modeling are 
tree-based models. For the particular case of circular data, the methodology 
of circular regression trees and forests is presented in the following section.

@@

\subsection{Data}
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airport for Innsbruck and Vienna on an
hourly temporal resolution. As predictor variables, we use temporal information
and $1$-hourly resolved $10$\,min mean observations of various meteorological quantities such as 
wind direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within an area of approximately 30\,km$^2$ are used. A topographical overview of
the airports and their surrounding areas with the station sites employed in
this study is provided in Figure~XXX. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

@@

    using the same daytime of $31$~days centered around the day of interest over all available years in the training data.

@@

    the regime changes during the forecast

@@

    maximum of $50$ classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}).

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular regression tree and the persistency model; except for the $3$\,h~forecasts
at Innsbruck where the persistency model is outperformed by all others. 
In comparison to the circular regression tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores of wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains only a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift, 
e.g., from up-valley to down-valley. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a change of the wind regime seem to 
lead to a higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

@@

\section{Tree algorithm}
\label{app:algorithm}
This section provides a more detailed overview on the permutation-test-based CTree algorithm 
\citep{Hothorn+Hornik+Zeileis:2006}, 
specifically for circular data as applied for building circular regression trees and forests 
as presented in this case study. 
An alternative tree-building framework is provided by the MOB algorithm, which is based on 
M-fluctuation tests \citep[see][for more details]{Zeileis+Hothorn+Hornik:2008}.

@@

To map the observed multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
\begin{equation}
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
\end{equation}
where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $T_l$,
as derived by \cite{Strasser+Weber:1999} and used to standardize $t_l$, and $\Sigma_l^+$ 
is the Moore-Penrose inverse of $\Sigma_l$.
As an alternative, also a maximum form ($c_{\text{max}}$) can be considered such that
the maximum of the absolute values of the standardized linear statistic is returned.
